Incremental tactics

### 2012-05-29 doc/papers/LFMTP2012/paper.tex
paper

The positive impact of rich type systems on programming language
design is not to be demonstrated anymore, both in terms of the safety
of programs --- they ``do not go wrong'', and programming practices
--- modularity, early bug catching. Verifying well-typing of
real-world programs is thus a critical matter, but it is a complex
operation, for the compiler designer and for the human operator.

@@

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (``type inference''), in which case the
missing information has to be algorithmically synthesized, but is also
true in many cases for what is commonly referred as ``type checking'',
or explicit ``church-style'' calculi. Take for example dependent type
systems, relying on a certain notion of definitional equality on
types, or type systems with subtype polymorphism: both rely on a rule
of the form
$$
\infer{\Gamma\vdash M : A \\ A\leq B}{\Gamma\vdash M : B}\ \text{,}
$$
for a given notion of $\leq$, which is not syntax-directed. The
implementer then has to convert the declarative system to an
algorithm, and convince himself that both are equivalent. This task,
often neglected, is nonetheless highly non-trivial and non-modular.

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to ``forget'' about the declarative system and take only the
result of the algorithm into consideration. But for the sake of the
discussion, let us choose another path: derivations in the declarative
systems, if they cannot always be easily inferred, can be
\emph{verified} if provided by an oracle. Even if the oracle is wrong,
we can tell cheaply by verifying the witness typing derivation it
provides. Let us then imagine a type-checker to be the pair of 1/ an
untrusted procedure $\finfer M$ to infer a typing derivations out of
programs $M$; it doesn't have to be complete, or even sound, and 2/ a
trusted and fast kernel $\fcheck{\md}$ to verify these derivations in
a given set of inference rules $\Sigma$. This is a \emph{certifying}
scheme: if $\fcheck{\finfer M} = \cst{true}$ then there really
\emph{is} a derivation $\md$ for $M$. Of course the converse is not
necessarily true: function $\finfer{}\!\!$ could fail, or have bugs
(return an incorrect derivation). Explicitly generating the derivation
in the declarative system has the following advantages over the first
approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\item for optimization purpose, compilers often need to propagate type
  information in the chain of transformation; the typing derivation
  produced by $\finfer{}$ being explicit, it could constitute the data
  structure being exploited.
\end{itemize}
\ldots\ but there is more:

@@

As type systems become richer, it becomes increasingly difficult for
the programmer to write a correct program in one try, compile it and
run it: writing a program longer than a few dozen of lines becomes a
tight and non-linear interaction between the human and the
type-checker involving constant experiments, fixes, etc. If this
interaction is possible for short programs and fast checkers, it
becomes increasingly hindered by the latency of complete rechecks:
even if the modifications made to the program between two interactions
are small, the whole program is usually rechecked. This is especially
true for languages embedding formal verification aspects, like proof
assistants: there, tight interaction with a type or proof checker is
unavoidable due to the difficulty of writing correct proofs without
guidance from the system, and the time taken by many tools to infer
easy parts of the proofs (proof search). The constant modification of
the source makes it necessary to observe \emph{incrementally} the
effect of a small change on the whole edifice: we need to take
advantage of the results of previous checks in order to build the
result for a modified version of the program.

Several methods have been devised for turning a batch procedure into
an incremental one \cite{pugh1989incremental,
  acar2003selective}. Unfortunately, they all suffer from their
genericity on the particular problem of type checking:
\begin{itemize}
\item first, they ignore the higher-order nature of program syntax,
  and the kind of structural quotients we make on syntax trees with
  binders and environments ($\alpha$-equivalence, weakening\ldots): a
  ground derivation of $\vdash 2+2 : \cst{nat}$ is still valid when
  put under a binder $\vdash\lam x 2 + 2 - x:\cst{nat}\to\cst{nat}$
  and should not be regenerated;
\item for very large programs, we might want to feed the type checker
  only with a small \emph{delta} on the previous version and not the
  whole program, thus saving the time to recognize already checked
  parts;
\item finally, there is no way to \emph{a posteriori} validate the
  result of the incremental process: the user has to trust not only
  the type-checking algorithm but also its transformation into an
  incremental one.
\end{itemize}
It just so happens that, if we consider as above a type-checker as a
derivation-generating procedure, and storing these derivations between
calls to the procedure, we could recognize already-built
subderivations for subterms common to a checked and a new version to
check of the program, and graft these pieces of derivation to form a
new, large one for the whole modified program, saving the regeneration
of all equal subterms' derivations in compatible contexts. This way we
maintain the verifiable witness of well-typing for the program, and
take advantage of the quotients made on higher-order syntax.

@@

In this paper, we present a framework to incrementally generate typing
certificates represented in the \LF\ logical framework, and implemented
as an \sysname{ML} library. We specifically propose:
\begin{itemize}
\item a simple language for representing programs and proof
  deltas, derived from contextual \LF\ \cite{nanevski2008contextual};
\item an untyped, unsafe calculus to write derivation-generating type
  checkers, based on a notion of function inverse %TODO le nom?
  to refer an already-computed value;
\item a generic algorithm to evaluate these type-checkers on program
  deltas and verify on-the-fly the generated certificates
  incrementally against a \emph{repository} of already constructed and
  checked derivations.
\end{itemize}

@@

A student is learning about System \sysname T, and wants to write a
program computing $(2+2)\times 3$. He knows that sometimes those
$\cst{rec}$ arguments are quite hard to get correct, but System
\sysname T is a typed language; he wants to take advantage of this to
build his program incrementally using the teacher's provided
incremental verifier. As a first attempt, he wants to verify $P_1$:
$$
\recb{\s\z}{\s\z}{\lam x\lam y \s {\var x}}\ \text{.}
$$
For this, he sends the command $\finfer{P_1}$ to the system. The
$\finfer{}$ operator takes a well-formed program and tries to construct
a valid typing derivation for its argument in the current, empty
context. The system succeeds and returns a typing derivation $\md_1$
of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} {\lam x \lam y \s{\var x}}}
  \\
  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function $\lam x\lam y\s{\var
  x}$, and he could reuse the pieces of derivation just built. Let us
call these $\md_2$, $\md_3$ and $\md_4$. Provided there is an operator
$\fget\md$ from the derivation to the program it types, what needs
really to be sent to the system is:
\begin{align*}
  &\finfer{(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} {\fget{\md_4}}}
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )}
\ \text{,}
\end{align*}
that is, only the subterms of the program that changed. Derivations
$\md_2$ and $\md_4$ used to live in the empty context, we are then
free to use it in the new context where
$\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T enjoys
the \emph{weakening} property. Taking advantage of $\md_2$ and $\md_4$
the system succeeds with a new derivation made out of these, without
having to regenerate them because it knows that $\finfer{\fget\md} =
\md$. Let us call $\md_5$ the subderivation generated for $\tlam x
{\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y}
{\fget{\md_4}}$, $\md_6$ for the arguments
$\app{\fget{\md_2}}{\fget{\md_2}}$ and $\md_7$ for the call
$\app{\var{add}} {\fget{\md_6}}$. Even if it occurs originally in the
scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
\emph{strengthening} property. Contrarily, $\md_7$ does: each time we
want to reuse it, we will have to provide a proof of this fact.

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \finfer{}(
  &\letb{add}{\fget{\md_5}}
  \\
  &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\fget{\gsubst{\md_7}{\msubst H {\finfer{\var{add}}}}})}
  {(\s{\fget{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all derivations for equal
subterms. In particular, $\md_7$ expects a proof that the variable
$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
$\finfer{\var{add}}$ meaning: at this point we will have generated (or
reused in this case) the derivation of $\var{add}$, so just use it.

Finally, he decides (in an odd move) to inline the definition of
$\var{mul}$ into its call site. Call $\md_8$ the derivation of the
definition of $\var{mul}$ (it depends on an $\var H$ of
$\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
$(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
$\times$. He orders:
\begin{align*}
\finfer{}(
\letb{add}{\fget{\md_5}}
\fget{({\gsubst{\md_9} {\msubst {H} {(\fget{\var{add}})}; \msubst {H'}
      {\gsubst{\md_8}{\msubst H {(\fget{\var{add}})}}}}})}
)
  \ \text{.}
\end{align*}
The definition for $\var{add}$ is unchanged, but the instantiation of
open derivations enables to directly substitute one in the other and
get the appropriate program: call $\md_{10}$ the whole derivation just
returned, $\fget{\md_{10}}$ returns:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} {\lam x \lam y \s{\var x}}}
  \\&\quad
  {\tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
      \app{\var{add}}\app{\var x} {\var z}}}
  \\&\quad\quad
  \app{(\app {\var{add}} \app {(\s{\s\z})}
    {(\s{\s\z})})}{(\s{\s{\s\z}})}
  \ \text{.}
\end{align*}

@@

A memoized type-checker stores in a table bindings from its input (the
program) to its output (the derivation), and returns automatically the
stored output if the current input matches a binding in the table. It
thus performs two operations at once: output reuse and recognition of
already-seen input. Since       %TODO

To simplify the problem,
we separate the task in:
\begin{itemize}
\item from a modified program, generate the \emph{delta} against the
  previous version, that is, observing and linking to unchanged
  derivations.
\item given a \emph{delta}, construct a new derivation and verify it.
\end{itemize}
The latter is the subject of this paper.

@@

\subsection{The \LF\ representation language}

The emission of certificates by untrusted applications first raises
the question of the \emph{data structure} to use for these
certificates. The requirements are:
\begin{itemize}
\item conciseness of the certificates, since they are usually stored
  or transferred remotely; in particular,
\item a well-understood, short and independent verification algorithm;
  this is known as the \emph{de Bruijn principle}.
\item universality with respect to the particular problem: having a
  certificate language for each domain of application would just shift
  the problem of trust; having one universally accepted certificate
  language that can be tailored to many problems puts a much stronger
  trust on the unique verifier.
\end{itemize}

The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is a well-accepted solution to this question. It is a typed
lambda-calculus and can be viewed as a language for
\emph{representing} terms, derivations and statements of $\Pi_0^1$
meta-theorems (but not their proofs) in a given inference system
supporting hypothetical judgments \cite{pfenning2001logical}. For
these, it provides built-in support for encoding languages with
binders or hypothesis (a feature called $\lambda$-tree syntax or HOAS
\cite{pfenning1988higher}). The user gives a \emph{signature} defining
abstract syntax and inference rules of his \emph{object language}, and
gets an algorithm to verify terms and derivations expressed in the
\LF\ syntax.

This framework has been used to issue certificates for a number of
practical applications like proof-carrying code
\cite{necula1997proof}, authentication framework \cite{appel1999proof}
\ldots %TODO ATP

In all these applications, the user feeds the generation program with
a query (the program to compile to machine-code, the access query, the
statement of a theorem \ldots) and gets in return a result (machine
code, yes/no\ldots) together with a certificate that the result
verifies the wanted property (semantics preservation, justification
for granting access, proof of the theorem\ldots). A potential client
can then verify independently that this result is adequate given the
query: it suffices to batch-verify the certificate.

\subsection{Proof-carrying static analysis}

\subsection{Storing computation results}
\subsection{Naming intermediate results}

\section{Examples}

@@

In this variant, only $\eta$-long $\beta$-normal forms have an
existence: canonicity for $\beta$-reduction is enforced in the syntax
and canonicity for $\eta$-expansion will be enforced by typing,
following \cite{hl07mechanizing}. Besides, application of a
\emph{head} (\ie\ a constant or variable) is $n$-ary (\ie\ to a
\emph{spine} of objects) as opposed to the more standard binary
application: this eases the definition substitution and
$\eta$-expansion, and corresponds to a focused sequent calculus known
as \sysname{LJT} \cite{herbelin1995λ}.

>>>>

The positive impact of rich type systems on programming language
design is not to be demonstrated anymore, both in terms of the safety
of programs --- they ``do not go wrong'', and programming practices
--- modularity, early bug catching. Verifying well-typing of
real-world programs is thus a critical matter, but it is a complex
operation, for the compiler designer and for the human operator:

@@

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (``type inference''), in which case the
missing information has to be algorithmically synthesized, but is also
true in many cases for what is commonly referred as ``type checking'',
or explicit ``church-style'' calculi. Take for example dependent type
systems, relying on a certain notion of definitional equality on
types, or type systems with subtype polymorphism: both rely on a rule
of the form
$$
\infer{\Gamma\vdash M : A \\ A\leq B}{\Gamma\vdash M : B}
$$
(for a given notion of $\leq$) which is not syntax-directed. The
implementer then has to convert the declarative system to an
algorithm, and convince himself that both are equivalent. This task is
highly non-trivial and non-modular: for instance, how to infer the
type of a System \sysname{T} recursor
$$
\infer{
  \Gamma\vdash M : \cst{nat}
  \and
  \Gamma\vdash N : A
  \and
  \Gamma, \var x:\cst{nat}, \var y:A\vdash P : A
}{
  \Gamma\vdash \recb M N x y P
}
$$
in presence of subtyping? The problem is that we have to guess a type
$A$ more general than that of $N$ and $P$, while giving that type to
variable $\var y$. A well-known algorithm if the subtyping relation
forms a complete lattice is to iterate the typing of $P$ with the join
of the previously computed and the new type until a fixpoint is
reached.

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to take only the result of the algorithm into
consideration. But let us choose another path for a minute:
derivations in the declarative systems, if they cannot always be
easily inferred, can be \emph{verified} if provided by an oracle. If
the oracle is wrong, we can tell cheaply by verifying the witness
typing derivation it provides. Let us then imagine a type-checker to
be the pair of 1/ an untrusted procedure $\finfer M$ to infer a typing
derivation out of program $M$; it doesn't have to be complete, or even
sound, and 2/ a trusted and fast kernel $\fcheck{\md}$ to verify these
derivations in a given set of inference rules $\Sigma$. This is a
\emph{certifying} scheme: %TODO cite Leroy
if $\fcheck{\finfer M} = \cst{true}$ then there really \emph{is} a
derivation $\md$ for $M$, but the converse is not necessarily true:
function $\finfer{}\!\!$ could fail, or have bugs (return an incorrect
derivation). Explicitly generating the derivation in the declarative
system has the following advantages over the first approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item compilers often need to propagate type information further in
  the chain of transformation; the typing derivation produced by
  $\finfer{}$ being explicit, it could constitute the data structure
  being exploited.
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\end{itemize}
\ldots\ but there is more:

@@

As type systems become richer, it becomes increasingly difficult for
the programmer to write a correct program in one try, compile it and
run it: writing a program longer than a few dozens of lines becomes a
tight and non-linear interaction between the human and the
type-checker involving constant experiments, fixes, etc. If this
interaction is possible for short programs and fast checkers, it
becomes increasingly hindered by the latency of complete rechecks:
even if the modifications made to the program between two interactions
are small, the whole program is usually rechecked. This is especially
true for languages embedding formal verification aspects, like proof
assistants: there, tight interaction with a type or proof checker is
unavoidable due to the difficulty of writing correct proofs without
guidance from the system, and the time taken by many tools to infer
easy parts of the proofs (proof search). The constant modification of
the source makes it necessary to observe \emph{incrementally} the
effect of a small change on the whole edifice: we need to take
advantage of the results of previous checks in order to build the
result for a modified version of the program.

Several methods have been devised for turning a batch procedure into
an incremental one \cite{pugh1989incremental,
  acar2003selective}. Unfortunately, they all suffer from their
genericity on the particular problem of type checking:
\begin{itemize}
\item first, they ignore the higher-order nature of program syntax,
  and the kind of structural quotients we make on syntax trees with
  binders and environments ($\alpha$-equivalence, weakening\ldots): a
  hypothetical derivation of\ \ $\vdash x+2 : \cst{nat}$ is still valid
  when put under a binder\ \ $\vdash\lam y (x + 2) -
  y:\cst{nat}\to\cst{nat}$ and should not be regenerated;
\item for very large programs, we might want to feed the type checker
  only with a small \emph{delta} on the previous version and not the
  whole program, thus saving the time to recognize already checked
  parts;
\item finally, there is no way to \emph{a posteriori} validate the
  result of the incremental process: the user has to trust not only
  the type-checking algorithm but also its transformation into an
  incremental one, which increases the size of the trusted base.
\end{itemize}

Now, if we consider as above a type-checker as a derivation-generating
procedure, and storing these derivations between calls to the
procedure, we could recognize already-built subderivations for
subterms common to a checked and a new version of the program, and
graft these pieces of derivation to form a new, large one for the
whole modified program, saving the regeneration of all equal subterms'
derivations in compatible contexts. This way we constantly maintain
the verifiable witness of well-typing for the program, and take
advantage of the quotients made on higher-order syntax.

@@

In this paper, we present a framework to incrementally generate typing
certificates represented in the \LF\ logical framework, and implemented
as an \sysname{ML} library. We specifically propose:
\begin{itemize}
\item a simple language for representing programs and proof
  deltas, derived from contextual \LF\ \cite{nanevski2008contextual};
\item a computational language to write untrusted
  derivation-generating type checkers, based on a notion of function
  inverse %TODO le nom?
  to refer to an already-computed value;
\item a generic algorithm to evaluate these type-checkers on program
  deltas and verify on-the-fly the generated certificates
  incrementally against a \emph{repository} of already constructed and
  checked derivations.
\end{itemize}

@@

A student is learning about System \sysname T --- a simply typed
$\lambda$-calculus with natural numbers introduction $\z$ and $\s{M}$,
and elimination ($\recb M N x y P$).  He (or she) wants to write a
program computing $(2+2)\times 3$. He knows that sometimes those
$\cst{rec}$ arguments are quite hard to get correct, but System
\sysname T is a typed language; he wants to take advantage of this to
build his program incrementally using the teacher's provided
incremental verifier. As a first attempt, he wants to verify $P_1$:
$$
\recb{\s\z}{\s\z} x y {\s {\var x}}\ \text{.}
$$
For this, he sends the command $\finfer{P_1}$ to the system. The
$\finfer{}$ operator takes a well-formed program and tries to construct
a valid typing derivation for its argument in the current, empty
context. The system succeeds and returns a typing derivation $\md_1$
of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} u v {\s{\var u}}}
  \\
  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function body $\s{\var x}$, and
he could reuse the pieces of derivation just built. Let us call these
respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
depends on the hypothesis $H$ that\ \ $\vdash \var x:\cst{nat}$ .
Provided there is an operator $\fget\md$ from the derivation to the
program it types, what needs really to be sent to the system is:
\begin{align*}
  &\finfer{}(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
    \\&\qquad
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )
\ \text{,}
\end{align*}
that is, only the changed subterms of the program. Derivations $\md_2$
used to live in the empty context, we are then free to use it in the
new context where $\var{add}:\cst{nat}\to\cst{nat}$, because System
\sysname T enjoys the \emph{weakening} property; $\md_4$ depending on
a derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole
with the derivation $\finfer{\var u}$ inferred recursively for then
new recursive argument $\var u$. Taking advantage of $\md_2$ and
$\md_4$ the system succeeds with a new derivation made out of these,
without having to regenerate them because it knows that
$\finfer{\fget\md} = \md$.

Let us call $\md_5$ the subderivation generated for $\tlam x
{\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y} u v
{\fget{\gsubst{\md_4}{\msubst H {\finfer{\var u}}}}}$, $\md_6$ for the
arguments $\app{\fget{\md_2}}{\fget{\md_2}}$ and $\md_7$ for the call
$\app{\var{add}} {\fget{\md_6}}$. Even if it occurs originally in the
scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
\emph{strengthening} property. Contrarily, $\md_7$ does: each time we
want to reuse it, we will have to provide a proof of this fact.

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \finfer{}(
  &\letb{add}{\fget{\md_5}}
  \\
  &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} z t {
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\fget{\gsubst{\md_7}{\msubst H {\finfer{\var{add}}}}})}
  {(\s{\fget{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all derivations for equal
subterms. In particular, $\md_7$ expects a proof that the variable
$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
$\finfer{\var{add}}$ meaning: at this point we will have generated (or
reused in this case) the derivation of $\var{add}$, so we just have to
use it here.

Finally, he decides (in an odd move) to inline the definition of
$\var{mul}$ into its call site. Call $\md_8$ the derivation of the
definition of $\var{mul}$ (it depends on an $\var H$ of
$\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
$(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
$\times$. He orders:
\begin{align*}
\finfer{}(
\letb{add}{\fget{\md_5}}
\fget{({\gsubst{\md_9} {\msubst {H} {(\fget{\var{add}})}; \msubst {H'}
      {\gsubst{\md_8}{\msubst H {(\fget{\var{add}})}}}}})}
)
  \ \text{.}
\end{align*}
The definition for $\var{add}$ is unchanged, but the instantiation of
open derivations enables to directly substitute one in the other and
get the appropriate program: call $\md_{10}$ the whole derivation just
returned, $\fget{\md_{10}}$ returns:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} x y {\s{\var x}}}
  \\&\quad
  {\tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} z t {
      \app{\var{add}}\app{\var x} {\var z}}}
  \\&\quad\quad
  \app{(\app {\var{add}} \app {(\s{\s\z})}
    {(\s{\s\z})})}{(\s{\s{\s\z}})}
  \ \text{.}
\end{align*}

@@

A memoized type-checker stores in a table bindings from its input (the
program) to its output (the derivation), and returns automatically the
stored output if the current input matches a binding in the table. It
thus performs two operations at once: output reuse and recognition of
already-seen input. Recognition --- that is from a modified program,
generate a delta from previous versions --- is in the higher-order
case a complex task that we hope to address in the future; we tackle
here the verification of output reuse: given a delta, construct a new
derivation and verify it.

@@

\subsection{The representation language}

The emission of certificates by untrusted applications first raises
the question of the \emph{data structure} to use for these
certificates. The requirements are:
\begin{itemize}
\item conciseness of the certificates, since they are stored or even
  transferred remotely;
\item a well-understood, short and independent verification algorithm.
  This is known as the \emph{de Bruijn principle};
\item universality with respect to the particular problem: having a
  certificate language for each domain of application would just shift
  the problem of trust; having one universally accepted certificate
  language that can be tailored to many problems puts a much stronger
  trust on the unique verifier.
\end{itemize}

The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is a well-accepted solution to this question. It is a typed
lambda-calculus and can be viewed as a language for
\emph{representing} terms, derivations and statements of $\Pi_0^1$
meta-theorems (but not their proofs) in a given inference system
supporting hypothetical judgments \cite{pfenning2001logical}. For
these, it provides built-in support for encoding languages with
binders and hypothesis (a feature called $\lambda$-tree syntax or HOAS
\cite{pfenning1988higher}). The user gives a \emph{signature} defining
abstract syntax and inference rules of his \emph{object language}, and
gets an algorithm to verify terms and derivations expressed in the
\LF\ syntax. This framework has been used to issue certificates for a
number of practical applications like proof-carrying code
\cite{necula1997proof}, authentication framework \cite{appel1999proof}
\ldots %TODO ATP

To write deltas by sharing subterms, we need to be able to address any
subterm and instantiate its free variables. For this, we define the
the \emph{slicing} of a well-typed \LF\ term: we extend \LF\ with
metavariables $\meta X$, $\meta Y$\ldots\ as in
\cite{nanevski2008contextual}. Metavariables stand for well-typed
terms defined in a repository, and are instantiated wrt. their free
variables. A repository also contains a distinguished metavariable
corresponding to the head of the term stored in it. For instance, the
repository
$$
\left(\begin{array}{llll}
    \meta X &\mapsto
    \envnil&\vdash
    \app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} &:
    \cst{tm}
    \\
    \meta Y &\mapsto
    \env u {\cst{tm}} &\vdash
    \app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
      {\var u}} v {\var y} } &:
    \cst{tm}
    \\
    \meta Z &\mapsto
    \envcons {\env u {\cst{tm}}} v {\cst{tm}} &\vdash
    \app{\app{\cst{app}}{\var v}}{\var u} &:
    \cst{tm}
    \\
    \meta T &\mapsto
    \env x {\cst{tm}} &\vdash
    \app{\cst{app}}\app{\var x}{\cst{0}} &:
    \cst{tm}

\end{array}\right), \smeta X {}
$$
is one of the possible way of slicing of the term
$$
\fbox{$\app {\cst{lam}} \lam x
\fbox{$\app {\cst{lam}} \lam y
\fbox{\app {\cst{app}} \app {\var y}
}$}
{(\fbox{\app
  {\cst{app}}\app{\var x}{\cst 0}})
}
$}$$
where all boxed subterms are referable slices.

\subsection{Computational calculus}
\subsection{Slicing results of computation}

@@

In this variant, only $\eta$-long and $\beta$-normal values have an
existence: canonicity for $\beta$-reduction is enforced in the syntax
and canonicity for $\eta$-expansion will be enforced by typing,
following \cite{hl07mechanizing}. Besides, application of a
\emph{head} (\ie\ a constant or variable) is $n$-ary (\ie\ to a
\emph{spine} of objects) as opposed to the more standard binary
application: this eases the definition substitution and
$\eta$-expansion, and corresponds to a focused sequent calculus known
as \sysname{LJT} \cite{herbelin1995λ}.
